### 2024-04-11 docs/SOSP 2024/hir-mir-opts.tex
Edit schedule and hir-mir section

The \Treebeard{} infrastructure was originally designed to target CPUs \cite{Treebeard}. 
However, it implements several optimizations on the HIR and MIR that 
can be leveraged across target processors and we find that these 
are beneficial for GPUs as well. This reuse of optimizations is possible 
because the intermediate representations on which these optimizations are performed 
are abstract and are designed to be target-independent. We briefly review theese 
optimizations below.

\subsection{Optimizations on High-Level IR}
We augment the existing \Treebeard{} infrastructure with loop rewrites on 
the HIR that are implemented through the scheduling language (Section \ref{sec:schedule}).
We use these to implement the automatic scheduling described in Section \ref{sec:exploring}. 
Additionally, the \Treebeard{} infrastructure implements HIR transformations to reorder and pad 
trees. It also implements tree tiling transformations on the HIR \cite{Treebeard}. We reuse the 
reordering and padding transformations on the HIR for GPUs. However, we found that 
tiling trees was not beneficial for GPUs. This is because the tiling transformations
introduces redundant computation inorder to vectorize computation on CPUs where 
all lanes need to follow the same control flow. However, on SIMT GPUs, we find that 
the benefits of tiling (coalescing memory accesses) do not outweigh the cost of
redundant computation. We leave an investigation of this for future work.

\subsection{Optimizations on Mid-Level IR}
The original \Treebeard{} infrastructure implements optimizations like 
tree-walk unrolling, tree-walk interleaving, and parallelization
on the MIR. These optimizations are beneficial for GPUs as well.
and the design of \Treebeard{} allows us to reuse the tree-walk 
unrolling and tree-walk interleaving optimizations on the MIR for GPUs.
\TODO{Should we talk about how interleaving is implemented as a 
statemachine and therefore it can be used across representations 
and tile traversal techniques?}

\subsection{A Note on Low-level IR}
Some changes to the original \Treebeard{} design were required to get
LIR to correctly lower to GPU code. The most important of these was 
the change to how the compiler implements support for in-memory 
representations of models (Section \ref{sec:representations}).
With these design changes, we were able to reuse much of the CPU 
implementation while customizing some parts for GPUs (for example,
buffers need to be allocated differently for CPU and GPU, caching 
is implemented differently etc.).

>>>>

The \TreebeardOLD{} infrastructure was originally designed to target CPUs \cite{Treebeard}. 
However, it implements several optimizations on the HIR and MIR that 
can be leveraged across target processors and we find that some these 
are beneficial for GPUs as well. This reuse of optimizations is possible 
because the intermediate representations on which these optimizations are performed 
are abstract and are designed to be target-independent. We briefly review these 
optimizations below.

\subsection{Optimizations on High-Level IR}
We augment the existing \TreebeardOLD{} infrastructure with loop rewrites on 
the HIR that are implemented through the scheduling language (Section \ref{sec:schedule}).
We use these to implement the automatic scheduling described in Section \ref{sec:exploring}. 
Additionally, the \TreebeardOLD{} infrastructure implements HIR transformations to reorder and pad 
trees. It also implements tree tiling transformations on the HIR \cite{Treebeard}. We reuse the 
reordering and padding transformations on the HIR for GPUs. However, we found that 
tiling trees was not beneficial for GPUs. This is because the tiling transformation
introduces redundant computation inorder to vectorize computation on CPUs where 
all lanes need to follow the same control flow. However, on SIMT GPUs, we find that 
the benefits of tiling (coalescing memory accesses) do not outweigh the cost of
redundant computation. We leave an investigation of this for future work.

\subsection{Optimizations on Mid-Level IR}
The original \TreebeardOLD{} infrastructure implements optimizations like 
tree-walk unrolling, tree-walk interleaving, and parallelization
on the MIR. These optimizations are beneficial for GPUs as well.
and the design of \TreebeardOLD{} allows us to reuse the tree-walk 
unrolling and tree-walk interleaving optimizations on the MIR for GPUs.

While building \Treebeard{}, we found that one of the performance bottlenecks on the 
GPU was that warps spent significant time being stalled. Since GPUs 
implement scoreboarding~\cite{HennesseyPatterson}, we were able to alleviate this bottleneck by
interleaving tree walks. This significantly improved performance of generated 
code. We found it surprising that the use of ILP could benefit 
performance on the GPU.  
\TODO{Should we talk about how interleaving is implemented as a 
statemachine and therefore it can be used across representations 
and tile traversal techniques?}

\subsection{A Note on Low-level IR}
Significant changes to the original \TreebeardOLD{} design were required to get
LIR to correctly lower to GPU code. The most important of these was 
the change to how the compiler implements support for in-memory 
representations of models (Section \ref{sec:representations}).
With these design changes, we were able to reuse much of the CPU 
implementation while customizing some parts for GPUs (for example,
buffers need to be allocated differently for CPU and GPU, caching 
is implemented differently etc.).

### 2024-04-11 docs/SOSP 2024/schedule.tex
Edit schedule and hir-mir section

The default loop order (as currently generated by the compiler) is (batch, tree), i.e, for each row in the input batch, go over all trees.

\subsection{The XBoost Schedule}
XGBoost\cite{XGBoost} is a very popular gradient boosting library. 
It implements inference on the CPU by going over a fixed number of rows (64 in the previous version)
for every tree and then moving to the next tree. When all trees have been walked for this 
set of rows, the next set of rows is taken up. Also, different sets of rows are 
processed in parallel.

The schedule used by XGBoost can be represented in \Treebeard{}'s scheduling 
language as follows.
\begin{lstlisting}[style=c++]
  tile(batch, b0, b1, CHUNK_SIZE)
  reorder(b0, tree, b1)
  parallel(b0)
\end{lstlisting}

\subsection{Tahoe Schedules}
Tahoe\cite{Tahoe} has four strategies that it picks from for a given model. 
Each of these strategies can be encoded using \Treebeard{}'s scheduling language
as we show below. 
\begin{itemize}
  \item \textbf{Direct Method}: In this strategy, a single GPU thread walks all trees
  for a given input row. The schedule for this strategy is as follows.
\begin{lstlisting}[style=c++]
  tile(batch, b0, b1, ROWS_PER_TB)
  reorder(b0, b1, tree)
  gpuDimension(b0, grid.x)
  gpuDimension(b1, block.x)
\end{lstlisting}
  Here, \op{ROWS\_PER\_TB} is the number of rows that are processed by a single thread block.
  \item \textbf{Shared Data}: In this strategy, a thread block walks all the trees 
  for a given row in parallel. If threads walk multiple trees, each thread accumulates
  partial results. Finally, a thread block wide reduction is performed to compute 
  the prediction. The schedule for this strategy is as follows.
\begin{lstlisting}[style=c++]
  reorder(batch, tree)
  gpuDimension(batch, grid.x)
  gpuDimension(tree, block.x)
  cache(batch)
\end{lstlisting}
  \item \textbf{Shared Forest}: In this strategy, the whole model is loaded into 
  shared memory and subsequently, a single thread walks all trees for a particular
  row. The schedule for this strategy is as follows.
\begin{lstlisting}[style=c++]
  tile(batch, b0, b1, ROWS_PER_TB)
  tile(tree, t0, t1, N_TREES)
  reorder(b0, b1, t0, t1)
  cache(t0)
  gpuDimension(b0, grid.x)
  gpuDimension(b1, block.x)
\end{lstlisting}
  Here, we create a placeholder single iteration loop \op{t0} so that we can 
  specify that all trees are to be cached.
  \item \textbf{Shared Partial Forest}: In case the model is too large to fit into
  shared memory, the model is split into chunks and each chunk is loaded into shared
  memory. Again, as in the previous strategy, one thread walks all trees assigned to
  a thread block for a row. The schedule for this strategy is as follows.
\begin{lstlisting}[style=c++]
  tile(batch, b0, b1, ROWS_PER_TB)

  tile(tree, t0, t0Inner, TREES_PER_TB)
  tile(t0Inner, t1, t2, TREES_PER_TB)
  cache(t1)
  reorder(b0, t0, b1, t1, t2)

  gpuDimension(b0, grid.x)
  gpuDimension(t0, grid.y)
  gpuDimension(b1, block.x)
\end{lstlisting}
\end{itemize}

There are some simplifying assumptions and limitations in the current design
\begin{itemize}
  \item Tree traversals are considered atomic. There is no way to express partial
   tree traversals or schedule individual node/level computations. 
  \item Accumulation of tree predictions is done immediately (as opposed to, 
  for example, collecting all predictions and performing a reduction later).
\end{itemize}

>>>>

We now show how the implementation strategies of several existing systems can 
be encoded in \Treebeard{}'s scheduling language. First, we note that the 
default loop order is [\op{batch}, \op{tree}], i.e, for each row in the input
batch, go over all trees.

% \subsection{The XBoost Schedule}
XGBoost\cite{XGBoost} implements inference on the CPU by going 
over a fixed number of rows (64 in the previous version)
for every tree and then moving to the next tree. When all trees 
have been walked for this set of rows, the next set of rows is 
taken up. Also, different sets of rows are processed in parallel.
This schedule can be implemented in \Treebeard{}'s scheduling language 
as follows.
\begin{lstlisting}[style=c++]
  tile(batch, b0, b1, CHUNK_SIZE)
  reorder(b0, tree, b1)
  parallel(b0)
\end{lstlisting}

Tahoe\cite{Tahoe} has four strategies for inference on the GPU that it picks from for a given model. 
We show how two of these strategies can be encoded using \Treebeard{}'s scheduling language.
The rest can be encoded similarly. 
\begin{itemize}
  \item \textbf{Direct Method}: In this strategy, a single GPU thread walks all trees
  for a given input row. The schedule for this strategy is as follows.
\begin{lstlisting}[style=c++]
  tile(batch, b0, b1, ROWS_PER_TB)
  reorder(b0, b1, tree)
  gpuDimension(b0, grid.x)
  gpuDimension(b1, block.x)
\end{lstlisting}
  Here, \op{ROWS\_PER\_TB} is the number of rows that are processed by a single thread block.
  \item \textbf{Shared Data}: In this strategy, a thread block walks all the trees 
  for a given row in parallel. If threads walk multiple trees, each thread accumulates
  partial results. Finally, a thread block wide reduction is performed to compute 
  the prediction. The schedule for this strategy is as follows.
\begin{lstlisting}[style=c++]
  reorder(batch, tree)
  gpuDimension(batch, grid.x)
  gpuDimension(tree, block.x)
  cache(batch)
\end{lstlisting}
%   \item \textbf{Shared Forest}: In this strategy, the whole model is loaded into 
%   shared memory and subsequently, a single thread walks all trees for a particular
%   row. The schedule for this strategy is as follows.
% \begin{lstlisting}[style=c++]
%   tile(batch, b0, b1, ROWS_PER_TB)
%   tile(tree, t0, t1, N_TREES)
%   reorder(b0, b1, t0, t1)
%   cache(t0)
%   gpuDimension(b0, grid.x)
%   gpuDimension(b1, block.x)
% \end{lstlisting}
%   Here, we create a placeholder single iteration loop \op{t0} so that we can 
%   specify that all trees are to be cached.
%   \item \textbf{Shared Partial Forest}: In case the model is too large to fit into
%   shared memory, the model is split into chunks and each chunk is loaded into shared
%   memory. Again, as in the previous strategy, one thread walks all trees assigned to
%   a thread block for a row. The schedule for this strategy is as follows.
% \begin{lstlisting}[style=c++]
%   tile(batch, b0, b1, ROWS_PER_TB)

%   tile(tree, t0, t0Inner, TREES_PER_TB)
%   tile(t0Inner, t1, t2, TREES_PER_TB)
%   cache(t1)
%   reorder(b0, t0, b1, t1, t2)

%   gpuDimension(b0, grid.x)
%   gpuDimension(t0, grid.y)
%   gpuDimension(b1, block.x)
% \end{lstlisting}
\end{itemize}

There are some simplifying assumptions and limitations in the current design
of the scheduling language. Mainly, tree traversals are considered atomic and
accumulation of tree predictions is done immediately (as opposed to, 
for example, collecting all predictions and performing a reduction later).
However, we find that these are not significant limitations in practice as
the current design is able to express most strategies of interest.
